Introduction to Markov state modeling with the PyEMMA software --- v\versionnumber

### 2018-09-13 manuscript/manuscript.tex
minor changes

\newcommand{\versionnumber}{0.3}
\newcommand{\githubrepository}{\url{github.com/markovmodel/pyemma_tutorials}}

@@

In order to create a Markov state model for a dynamical system, each data point in the time series is assigned to a state.
Given an appropriate lag time, every pairwise transition at that lag time is counted and stored in a count matrix.
Then, the count matrix is converted to a row-stochastic transition probability matrix $P$, which is defined for the specified lag time.
For MD simulations in equilibrium, $P$ should obey detailed balance which is enforced by constraining the estimation of $P$ to the following equations:
\begin{equation}
\label{eq:balance}
\pi_i P_{ij} = \pi_j P_{ji},
\end{equation}
where $\pi_i$ is the stationary probability of state $i$ and $P_{ij}$ is the probability of transitioning to state $j$ conditional on being in state $i$.
The constraints (\ref{eq:balance}) are omitted if MD simulations are not conducted in equilibrium, e.g.,
for systems experiencing a pulling force or an external potential---see~\cite{Koltai2018} for a recent review on nonequilibrium MSMs.
For the remainder of this section we will simplify the matter by assuming the more common scenario of MD simulations without external forces and (\ref{eq:balance}) to hold.

@@

Once we have used the ITS to choose the lag time, we can check whether a given transition probability matrix $T(\tau)$ is approximately Markovian using the Chapman-Kolmogorov (CK) test~\cite{noe-folding-pathways}.
The CK property for a Markovian matrix is,
\begin{equation}
T(k \tau) = T^k(\tau),
\end{equation}
where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than~1, whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model.

@@

More details on how to model complex systems with the techniques presented here are described e.g.~by~\cite{plattner_protein_2015,plattner_complete_2017}.

>>>>

\newcommand{\versionnumber}{1.0}
\newcommand{\githubrepository}{\url{github.com/markovmodel/pyemma_tutorials}}

@@

In order to create a Markov state model for a dynamical system, each data point in the time series is assigned to a state.
Given an appropriate lag time, every pairwise transition at that lag time is counted and stored in a count matrix.
Then, the count matrix is converted to a row-stochastic transition probability matrix $P$, which is defined for the specified lag time.
For MD simulations in equilibrium, $P$ should obey detailed balance which is enforced by constraining the estimation of $P$ to the following equations:
\begin{equation}
\label{eq:balance}
\pi_i p_{ij} = \pi_j p_{ji},
\end{equation}
where $\pi_i$ is the stationary probability of state $i$ and $p_{ij}$ is the probability of transitioning to state $j$ conditional on being in state $i$.
The constraints (\ref{eq:balance}) are omitted if MD simulations are not conducted in equilibrium, e.g.,
for systems experiencing a pulling force or an external potential---see~\cite{Koltai2018} for a recent review on nonequilibrium MSMs.
For the remainder of this section we will simplify the matter by assuming the more common scenario of MD simulations without external forces and (\ref{eq:balance}) to hold.

@@

Once we have used the ITS to choose the lag time, we can check whether a given transition probability matrix $T(\tau)$ is approximately Markovian using the Chapman-Kolmogorov (CK) test~\cite{noe-folding-pathways}.
The CK property for a Markovian matrix is,
\begin{equation}
P(k \tau) = P^k(\tau),
\end{equation}
where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than~$1$, whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model.

@@

More details on how to model complex systems with the techniques presented here are described, e.g., by~\cite{plattner_protein_2015,plattner_complete_2017}.
